---
title: "Calculating-PPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating-PPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MarinePPI)
library(tidyverse)
library(sf)
```

To run MarinePPI, you will need:

* A viable **internet connection** so that MarinePPI subroutines can download rasters and shapefiles from GHSL,
* A data frame of **geospatial coordinates** specifying the marine site locations, and;
* To specify **buffer distance(s)** for PPI calculations. This is the radius that will be used for focal means and should be specified in kilometers (km). 

## Survey Sites
This vignette computes PPI for 22 sites in Puget Sound from Heery et al. (in preparation: *A standardized approach for quantifying urban-related losses in habitat-forming macroalgae and marine biogenic complexity*). The data frame containing lat/lons (in decimal degrees) is available as part of this package. A .csv of the file is also provided as a template: [here](<https://github.com/eheery/MarinePPI/blob/main/data/example_locations.csv>).

```{r }
# Convert to sf object
sites <- st_as_sf(PugetSoundSites, coords = c("Lon_dd", "Lat_dd"), crs = 4326)
sites
```


#### Geospatial coordinates

The csv of geospatial coordinates for this example is available [here](<https://github.com/eheery/MarinePPI/blob/main/data/example_locations.csv>). We import it into R and convert it to an [sf POINTS](<https://r-spatial.github.io/sf/articles/sf1.html>) object as follows:

